---
title: "GDPR Fines - Exploratory Data Analysis"
date: 2020-06-11
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    keep_md: yes
  pdf_document: default
bibliography: references.bib
csl: "https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl"
---

Data Privacy and Data Security -- two espoused concepts that gain substantial traction in the current times of Covid-19, especially in the context of the upcoming German Corona tracking app [@SAP.2020] that is about to be released next week. Public expectations compete with individual reservations and data privacy concerns, partly fueled by events such as the 2013 Snowden revelations or the 2018 Camebridge Analytica scandal. Likewise, events like these contributed to the willingness of the EU to enforce supranational regulations on data privacy and security, climaxing the in the EU-GDPR [@EuropeanUnion.2018] which is binding for all EU member states since 2018-05-25.

In this blog post, I would like to take the current debate as an opportunity to dive into a dataset about monetary fines since the issuance of the EU-GDPR. In the process of doing so, I will do some web scraping to extract the dataset from the *GDPR Fines Tracker* on *privacyaffairs.com* [@PrivacyAffairs.2020], clean the dataset using some of the tools recently released as part of the `dplyr v1.0.0` release [@Wickham.2020], experiment with some fuzzy string matching capabilities provided in the `stringdist` package [@vanderLoo.2014] and engage in some visual EDA. Meanwhile, this post is inspired by [David Robinson](https://www.youtube.com/watch?v=EVvnnWKO_4w)'s TidyTuesday Series. The whole analysis is embeeded into an R markdown document.

```{r setup, include = FALSE}
#clear environment
rm(list=ls())
#clear console
cat("\014")


knitr::opts_chunk$set(
  #set comment formatting
  comment = ">",
  #collapse code and output
  collapse = TRUE,
  #set standard figure size (0.618 as "golden" aspect ratio)
  fig.width = 6, fig.asp = 0.618,
  #set standard output size (i.e. arrangement of plot in markdown document; here: 70% of line width)
  out.width = "70%", fig.align = "center",
  #figure save options
  dev = "png", dpi = 300
)

#do once to import all available fonts
#font_import()
```

```{r package-management, warning = FALSE, message = FALSE}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rvest, tidyverse, kableExtra, extrafont, stringdist)
```

### Web Scraping

First, I extract the dataset from the *GDPR Fines Tracker* on *privacyaffairs.com* [@PrivacyAffairs.2020] using the `rvest` package and some regular expression (*regex*) to finally transform the JSON-format into a convenient data frame format.
```{r gdpr_data_raw}
gdpr_data <- read_html("https://www.privacyaffairs.com/gdpr-fines/") %>% 
  #find html_node that contains the data
  html_nodes(xpath = "(.//script)[1]") %>% 
  #extract text
  rvest::html_text() %>%
  #trim to produce clean json format
  str_sub(start = str_locate(., "\\[")[[1]], end = str_locate(., "\\]")[[1]]) %>% 
  #remove new lines and carriage returns
  str_remove_all(paste("\\n", "\\r", sep = "|")) %>% 
  #parse JSON
  jsonlite::fromJSON()
```

In total, `r dim(gdpr_data)[1]` violations are categorized by *PrivacyAffairs* as of the day of this blogpost (`r Sys.Date()`). I randomly sample five observations to get a first overview of the dataset.

```{r sample_table_1, echo = FALSE}
set.seed(27052020)

gdpr_data %>% 
  select(-summary, -picture) %>% 
  slice_sample(n = 5) %>% 
  knitr::kable(
    caption = "TEEEEEEEEESSSSST",
    format.args = list(big.mark = ',')
  ) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = T) %>% 
  row_spec(0, bold = TRUE, color = "white", background = "#8486B2")
```

### Data Cleaning

In a next step, I streamline the dataset a little more and try to get rid of the various inconsistencies in the data entries. 

First, I would like to adjust some of the column names to highlight the actual content of the features. I rename `name` to `country`, `controller` to `entity` (i.e. the entity fined as a result of the violation) and `articleViolated` to `violation`. Second, I infer from the sample above that the violation `date` is not in standard international date format (`jjjj-mm-dd`). Let's change this using the `lubridate` package while properly accounting for `r sum(gdpr_data$date == "01/01/1970")` `NA`s (indicated by unix time `1970-01-01`). Third, I format the `price` feature by specyfing it as a proper currency (using the `scales` package). Fourth, and analogue to the violation `date`, I properly account for `NA`s -- taking the form "Not disclosed", "Not available" and "Not known" -- in the `entity` as well as `type` feature (containing `r gdpr_data$controller %>% str_detect("Unknown|Not") %>% sum()` and `r gdpr_data$type %>% str_detect("Unknown|Not") %>% sum()` missing values, respectively). In addition, I must correct one errorneous fined entity (violation id `r gdpr_data %>% select(id, controller) %>% filter(controller == "https://datenschutz-hamburg.de/assets/pdf/28._Taetigkeitsbericht_Datenschutz_2019_HmbBfDI.pdf") %>% pull(id)`). Finally, I clean the `violation` predictor using regex and the `stringr` package.
```{r gdpr_data_clean}
gdpr_data <- gdpr_data %>% 
  #1
  rename(country = name, entity = controller, violation = articleViolated) %>% 
  #2
  mutate(across(date, ~na_if(lubridate::mdy(.), "1970-01-01"))) %>% 
  #3
  mutate(across(price, ~if_else(. == 0, NA_integer_, .))) %>% 
  #4
  mutate(across(c(entity, type), ~if_else(str_detect(., "Unknown|Not"), NA_character_, .))) %>% 
  mutate(across(entity, ~str_replace(., "https://datenschutz-hamburg.de/assets/pdf/28._Taetigkeitsbericht_Datenschutz_2019_HmbBfDI.pdf", "HVV GmbH"))) %>% 
  #5
  mutate(across(violation, ~if_else(is.na(str_extract(., ">.+<")), ., str_extract(., ">.+<") %>% str_sub(., 2, -2)))) %>%
  mutate(across(c(violation, authority), ~str_replace_all(., "\\t", "")))
```

Since a cross-check of the entity names reveals quite a few inconsistencies in how the entities have been written in the databse, I leverage the `stringdist` package [@vanderLoo.2014] for fuzzy string matching to homogenize some of the entries. For example, the *optimal string alignment* (*osa*) measure allows to assess the similarity of two strings by enumerating the number of pre-processing steps (deletion, insertion, substitution and transposition) necessary to transform one string into another. Adhering to the following assumptions yields four fuzzy matches which are accounted for in the subsequent EDA:
* Set the minimum-osa threshold to 3 (i.e. only consider string pairs which require three transformations to be aligned).
* Only consider strings of length > 3 (otherwise the minimum-osa threshold becomes redundant).
```{r fuzzy matching}
entities <- gdpr_data %>% 
  distinct(entity) %>% 
  drop_na %>% 
  mutate(id = row_number(), .before = 1)


fuzzy_matches <- unique(gdpr_data$entity[!is.na(gdpr_data$entity)]) %>% 
  expand_grid(ent_a = ., ent_b = .) %>% 
  mutate(osa = stringdist(ent_a, ent_b, method = "dl", nthread = 4)) %>% 
  filter(osa < 4L &
           osa != 0L &
           str_length(ent_a) > 3L &
           str_length(ent_b) > 3L) %>% 
  left_join(entities, by = c("ent_a" = "entity"), suffix = c(".a", ".b")) %>% 
  left_join(entities, by = c("ent_b" = "entity"), suffix = c(".a", ".b")) %>% 
  filter(id.a < id.b)

fuzzy_matches

gdpr_data <- gdpr_data %>% 
  mutate(across(entity, ~str_replace_all(., c("Telecommunication Service Provider" = "Telecommunication service provider",
                                              "A mayor" = "Mayor",
                                              "A bank" = "Bank",
                                              "Vodafone Espana" = "Vodafone España"))))
```

Finally, let's have a look at the cleaned data.
```{r sample_table_2, echo = FALSE}
set.seed(27052020)

gdpr_data %>% 
  select(-summary, -picture) %>% 
  slice_sample(n = 5) %>% 
  knitr::kable(
    caption = "TEEEEEEEEEEST",
    format.args = list(big.mark = ',')
  ) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = T) %>% 
  row_spec(0, bold = TRUE, color = "white", background = "#8486B2")
```

Now let's briefly validate the integrity of the scraped dataset.
```{r missing_values, echo = FALSE}
gdpr_data %>% 
  map_df(., ~sum(is.na(.))) %>% 
  knitr::kable(
    caption = "TEEEEST 2",
  ) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = T) %>% 
  row_spec(0, bold = TRUE, color = "white", background = "#8486B2")
```
And indeed, a quick glance at the missing values per feature reveals numerous missing values for the `price` (`r (gdpr_data %>% map_df(., ~sum(is.na(.))) %>% pull(price) / dim(gdpr_data)[[1]]) %>% scales::percent(accuracy = 0.01)`), `date` (`r (gdpr_data %>% map_df(., ~sum(is.na(.))) %>% pull(date) / dim(gdpr_data)[[1]]) %>% scales::percent(accuracy = 0.01)`) and `entity` (`r (gdpr_data %>% map_df(., ~sum(is.na(.))) %>% pull(entity) / dim(gdpr_data)[[1]]) %>% scales::percent(accuracy = 0.01)`) feature. Without diving deeper into the information sources, it may be assumed that for the affected court cases no complete record of the verdict was openly published by the jurisdiction.

Also, with regards to some of the fines, *PrivacyAffairs* explicitely states that "*The Marriott and British Airways cases are not final yet and the fines are just proposals. Other GDPR fines trackers incorrectly report those as final.*"

### Exploratory Data Analysis

```{r parameters, include=F}
#hier nochmal code einfügen
windowsFonts(gg_font = windowsFont("Garamond"))

font_size_title <- 14
font_size_subtitle <- 12
font_size_caption <- 10
font_size_other <- 10

signature_color <- "#8486B2"
```

First ever fine
```{r}
gdpr_data %>% 
  select(-summary, -picture) %>% 
  filter(date == min(date, na.rm = TRUE))
```

Top 10 fines
```{r}
gdpr_data %>%
  select(-summary, -picture) %>% 
  slice_max(order_by = price, n = 10) %>% 
  knitr::kable(
    caption = "TEEEEST 3",
  ) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = T) %>% 
  row_spec(0, bold = TRUE, color = "white", background = "#8486B2")
```

Top 10 fined entities
```{r top10}
euro <- scales::dollar_format(
  prefix = "",
  suffix = "",
  big.mark = ",",
  decimal.mark = "."
)

gdpr_data %>%
  drop_na(price, date) %>% 
  mutate_at(vars(entity), ~as.factor(.) %>% 
            fct_lump_n(., 8, w = price) %>% 
            fct_explicit_na(., na_level = "Other")) %>%
  
  
  ggplot(., aes(date, price / 1000000)) +
    scale_y_log10(labels = euro) +
    scale_x_date(date_breaks = "3 month",
                 date_minor_breaks = "1 month",
                 date_labels = "%Y-%m",
                 limits = c(as.Date("2018-05-01"), NA)) +
    geom_point(aes(color = entity)) +
    geom_point(size = 3, shape = 1, data = . %>% filter(entity != "Other")) +
    scale_color_brewer(palette = "Set3") +
    labs(
      x = "",
      y = "Imposed Fine [log m€]",
      title = "Distribution of Fines since 2018-05-28",
      subtitle = "Few very Heafty Fines Dominate the Landscape",
      caption = "Data from privacyaffairs.com"
    ) +
    theme_classic() +
    theme(
      text = element_text(family = "gg_font"),
      plot.title = element_text(size = font_size_title, face = "bold"),
      plot.subtitle = element_text(size = font_size_subtitle),
      plot.caption = element_text(size = font_size_caption, face = "italic"),
      axis.text = element_text(size = font_size_other, color = "black"),
      axis.title = element_text(size = font_size_other),
      legend.position = "bottom",
      legend.title = element_blank()) +
    guides(
      color = guide_legend(
        nrow = 3,
        override.aes = list(size = 4)
      )
    )
```

Art. 83 GDPR: 

https://gdpr.eu/article-83-conditions-for-imposing-administrative-fines/

Nr. 4: 10m€ or 2% of total revenue
Nr. 5: 20m€ or 4% of total revenue
```{r}
gdpr_data %>%
  drop_na(price) %>% 
  mutate_at(vars(entity), ~as.factor(.) %>% 
              fct_lump_n(., 8, w = price) %>% 
              fct_explicit_na(., na_level = "Other")) %>% 
  group_by(entity) %>% 
  summarise(total_fine = sum(price),
            fine_freq = n()) %>% 
  
  
  ggplot(., aes(total_fine / 1000000, fct_reorder(entity, total_fine))) +
    geom_col(fill = signature_color) +
    geom_text(aes(label = euro(total_fine / 1000000)),
              size = font_size_other * .35, family = "gg_font", fontface = "bold",
              nudge_x = 2.5) +
    geom_vline(xintercept = 10, linetype = "dashed") +
    geom_vline(xintercept = 20, linetype = "dashed") +
    scale_x_continuous(labels = euro, breaks = seq(0, 50, 10)) +
    labs(
      x = "Imposed Fine [m€]",
      y = "",
      title = "Top 8 Fined Entites",
      subtitle = "10m€ Threshold only Cracked by Five Individual Entities",
      caption = "Data from privacyaffairs.com"
    ) +
    theme_classic() +
    theme(
      text = element_text(family = "gg_font"),
      plot.title = element_text(size = font_size_title, face = "bold"),
      plot.subtitle = element_text(size = font_size_subtitle),
      plot.caption = element_text(size = font_size_caption, face = "italic"),
      axis.text = element_text(size = font_size_other, color = "black"),
      axis.title = element_text(size = font_size_other)
    )
```


```{r}
gdpr_data %>%
  drop_na(price) %>% 
  mutate_at(vars(country), ~as.factor(.) %>% 
              fct_lump_n(., 12, w = price) %>% 
              fct_explicit_na(., na_level = "Other")) %>% 
  group_by(country) %>% 
  summarise(total_fine = sum(price),
            fine_freq = n(),
            .groups = "rowwise") %>% 

  ggplot(aes(fct_reorder(str_wrap(country, 10), -total_fine), total_fine / 1000000)) +
    geom_col(fill = signature_color) +
    geom_text(aes(label = euro(total_fine / 1000000)),
              size = font_size_other * .35, family = "gg_font", fontface = "bold",
              nudge_y = 3) +
    scale_y_continuous(labels = euro, breaks = seq(0, 50, 10)) +
    labs(
      x = "Imposed Fine [m€]",
      y = "",
      title = "European Countries Ranked by Imposed Fines",
      subtitle = "The German 'Tin Foil' Lags Behind",
      caption = "Data from privacyaffairs.com"
    ) +
    theme_classic() +
    theme(
      text = element_text(family = "gg_font"),
      plot.title = element_text(size = font_size_title, face = "bold"),
      plot.subtitle = element_text(size = font_size_subtitle),
      plot.caption = element_text(size = font_size_caption, face = "italic"),
      axis.text = element_text(size = font_size_other, color = "black"),
      axis.text.x = element_text(angle = 90, hjust = 0.95,vjust = 0.2),
      axis.title = element_text(size = font_size_other)
    )
```

#which article was most frequently violated?
- drop all fines for which no GDPR article is stated
- on the country side we see a most diverse set of violated articles on the side of romania and hungary
- poland and croatia, almost no activity
```{r}


gdpr_data %>% 
  select(country, violation) %>% 
  separate_rows(violation, sep = ",") %>% 
  filter(str_detect(violation, "Art")) %>% 
  mutate(across(violation, ~str_extract(., "(?<=Art.\\s?)[0-9]+"))) %>%
  mutate(across(everything(), as.factor)) %>% 
  group_by(country, violation) %>% 
  summarise(freq = n(), .groups = "rowwise") %>% 
  
  
  ggplot(aes(
    x = fct_relevel(violation, function(x){as.character(sort(as.integer(x)))}),
    y = country
  )) +
    geom_point(aes(color = freq), shape = 15, size = 4) +
    scale_color_gradient(name = "Number\nof Fines", low = "#cdd8ff", high = "#021868") +
    labs(
      x = "GDPR Article Number",
      y = "",
      title = "Distribution of Violated Articles across Europe",
      subtitle = "Art. 5, 6, 32 GDPR as Major Stumbling Blocks"
    ) +
    theme_classic() +
    theme(
      text = element_text(family = "gg_font"),
      plot.title = element_text(size = font_size_title, face = "bold"),
      plot.subtitle = element_text(size = font_size_subtitle),
      plot.caption = element_text(size = font_size_caption, face = "italic"),
      axis.text = element_text(size = font_size_other, color = "black"),
      axis.title = element_text(size = font_size_other)
    )
```


Article that incurred the highes average fine
Compute average fine by dividng by the number of articles involved in the fine
```{r}
gdpr_data %>% 
  drop_na(price) %>% 
  select(violation, price) %>% 
  mutate(across(price, ~ . / str_count(violation, "Art"))) %>% 
  separate_rows(violation, sep = ",") %>% 
  filter(str_detect(violation, "Art")) %>% 
  mutate(across(violation, ~str_extract(., "(?<=Art.\\s?)[0-9]+") %>% as.factor %>% fct_lump(., 8, w = price))) %>%
  group_by(violation) %>% 
  summarise(avg_fine = sum(price),
            .groups = "rowwise") %>% 
  ungroup %>% 
  left_join(read_delim("./article_titles.txt", delim = "; ", col_types = cols("f", "c")), by = c("violation" = "article")) %>% 
  mutate(across(title, ~paste(violation, "-", .))) %>% 
  
  
  ggplot(aes(avg_fine / 1000000,
             title %>% str_trunc(50) %>% str_wrap(30) %>% as.factor %>% fct_reorder(., avg_fine, .desc = F))) +
    geom_col(fill = signature_color) +
    scale_x_continuous(labels = euro) +
    geom_vline(aes(xintercept = mean(avg_fine / 1000000)), linetype = "dashed") +
    labs(
      x = "Imposed Fine [m€]",
      y = "GDPR Article Number",
      title = "Average Fine per Violated Article",
      subtitle = "Art. 5, 6, 32 GDPR as Major Stumbling Blocks",
      caption = "Data from privacyaffairs.com"
    ) +
    theme_classic() +
    theme(
      text = element_text(family = "gg_font"),
      plot.title = element_text(size = font_size_title, face = "bold"),
      plot.subtitle = element_text(size = font_size_subtitle),
      plot.caption = element_text(size = font_size_caption, face = "italic"),
      axis.text = element_text(size = font_size_other, color = "black"),
      axis.title = element_text(size = font_size_other)
    )
```

```{r}
gdpr_data %>%
  drop_na(price, date) %>% 
  mutate(month = lubridate::month(date, label = T)) %>% 
  group_by(month) %>% 
  summarise(total_fine = sum(price),
            freq = n(),
            .groups = "rowwise") %>% 

  ggplot(aes(month, freq)) +
    geom_col(aes(fill = freq), width = 1, color = "white") +
    scale_y_continuous(limits = c(NA, 45)) +
    scale_fill_gradient(name = "Number\nof Fines", low = "#cdd8ff", high = "#021868")+
    coord_polar("x") +
    labs(
      x = "",
      y = "",
      title = "Yearly Ditribution of Fines",
      subtitle = "Hello Summer Slump - We're on Vacation",
      caption = "Data from privacyaffairs.com"
    ) +
    theme_classic() +
    theme(
      text = element_text(family = "gg_font"),
      plot.title = element_text(size = font_size_title, face = "bold"),
      plot.subtitle = element_text(size = font_size_subtitle),
      plot.caption = element_text(size = font_size_caption, face = "italic"),
      axis.text.x = element_text(size = font_size_caption),
      axis.text.y = element_blank(),
      axis.title = element_text(size = font_size_other),
      line = element_blank(),
      legend.position = "right"
    )
```

### References
